---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---

***
## Project Creation
***

Repository forked from rdpeng/RepData_PeerAssessment1 on "2015-02-10"

### Load some packages
```{r}
library(data.table)
library(lattice)
```

***
## Loading and preprocessing the data
***
###unzip file and verify file size

Load Data from provided file "activity.zip" to create activity.csv

```{r echo = TRUE}
unzip("activity.zip")

#verify file size

file.info("activity.csv")

```

### Load data into data.table "activity"

```{r echo = TRUE}

activity<-as.data.table(read.csv("activity.csv"))

#And check data loaded

head(activity)

```

For this section we will assume that days with NA values for activity$steps are to be removed.

```{r echo = TRUE}

activity<-activity[steps != "NA"]

#and check activity structure

head(activity)

```

***
## What is mean total number of steps taken per day?
***

### Calculate the total steps per day

We will use data.table functions to get total steps per day

```{r}
stepsPerDay <-activity[,sum(steps),by=date]

#And clean set the column name to something useful

setnames(stepsPerDay,"V1","TotalSteps")

names(stepsPerDay)

```

###Now make a histogram of the total steps per day


```{r}
h<-hist(stepsPerDay$TotalSteps,breaks=20,
        xlab="Total Steps per Day" ,main=""
      )

```

###Calculate mean and median total steps per day

We will now use the summary function to get the mean and median (along with a few other parameters)

```{r}
stats<-summary(stepsPerDay$TotalSteps)
```


The mean numbers of steps per data is `r stats["Mean"]` and the median is `r stats["Median"]`


***
## What is the average daily activity pattern?
***

### Create new averages by intervals

```{r}
stepsPerInterval <-activity[,mean(steps),by=interval]

#And clean set the column name to something useful

setnames(stepsPerInterval,"V1","MeanSteps")

names(stepsPerInterval)

```

### Now plots the results

```{r}
plot(stepsPerInterval$interval,stepsPerInterval$MeanSteps,
     xlab="5 min. Interval", ylab ="Mean Steps",type="l")

```

### Find interval with max. number of steps

```{r}

maxInterval<-stepsPerInterval$interval[
             which.max(stepsPerInterval$MeanSteps)]
```


The interval with the maximum number of averages steps if `r maxInterval`

***
## Imputing missing values
***

### lets start with a fresh copy of the data

activity<-as.data.table(read.csv("activity.csv"))

### create table of is.na results

```{r}
counts<-table(is.na(activity$steps))
```

there are `r counts[TRUE]` rows with NA values for the steps variable

We should check the other columns

```{r}
table(is.na(activity$date))

table(is.na(activity$interval))

```

Only the steps column has NA values

### Fill in NA values with the interval average

Here are the steps

1. create a subset with NA values "missing" and a subset that did have values "notMissing"
2. replace steps values in "missing" with means interval values
3. merge the new "missing"" with "notMissing"
4. sort to original order

```{r}
# step 1
missing<-activity[is.na(steps)]

notMissing<-activity[!is.na(steps)]

#step 2
missing<-merge(missing,stepsPerInterval,by ="interval")

#clean up columns

missing[, steps := NULL ]

setnames(missing,"MeanSteps","steps")

#Setp 3

imputed<-rbind(missing,notMissing)

#step 4

imputed<-imputed[order(imputed$date,imputed$interval)]


```

### now lets compare the results to the first section

Calculate new total steps per day

```{r}
istepsPerDay <-imputed[,sum(steps),by=date]

#And clean set the column name to something useful

setnames(istepsPerDay,"V1","TotalSteps")

names(istepsPerDay)

```

###Now calculate a histogram of the total steps per day but plot later.


```{r}
hi<-hist(istepsPerDay$TotalSteps,breaks=20,plot=FALSE)

```

###Calculate mean and median total steps per day

We will now use the summary function to get the mean and median (along with a few other parameters)

```{r}
istats<-summary(istepsPerDay$TotalSteps)
```

Let plot the histograms together

```{r}

plot(h,xlab="Total Steps",main ="With NA removed")
plot(hi,xlab="Total Steps", main ="Imputed")

```

Because we added steps we get a different histogram but we see the means and medians are the same.


Data Set   |  Mean                | Median
---------- |  --------------------| ---------------------
NA removed |  `r stats["Mean"]`   | `r stats["Median"]` 
Imputed    |  `r istats["Mean"] ` | `r istats["Median"]`



***
## Are there differences in activity patterns between weekdays and weekends?
***

###Starting with data.table "imputed"

### create a new column "weekday" that is TRUE for weekdays and FALSe for weekends

```{r}
imputed[, weekday := 
          as.factor(!weekdays(as.Date(imputed$date))
          %in% c("Sunday"  ,"Saturday"))
      ]


head(imputed)  

```


And now for the big plot

```{r}
intervalWeedayMeans<-imputed[,mean(steps),by=.(interval,weekday)]


xyplot(intervalWeedayMeans$V1~ intervalWeedayMeans$interval |
        intervalWeedayMeans$weekday, type="l",layout=c(1,2) ,
        xlab="interval" , ylab= "Mean Steps",
        strip=strip.custom(factor.levels=c("Weekend","Weekdays"))
       )



```


Created `r Sys.Date()` by Craig Parman

